Density-dependent selection and the limits of relative fitness

### 2018-03-27 manuscript.tex
l

In light of the complications of density-dependence, the assignment of a density-independent relative fitness to each type has been justified as an approximation that holds when selection is weak and $N$ changes slowly (\citealt{kimura1969natural}; \citealt[pp. 277]{ewens_2004}; \citealt[Chap. 4]{charlesworth_1994}). Under these conditions, $s$ is approximately constant in Eq.~\eqref{eq:canonical}, at least for some number of generations. If $s$ depends only on density, not frequency, this approximate constancy can hold over entire selective sweeps \citep[Fig. 9.5]{otto_2011}.

@@

Here we argue that the simplified models of density-regulated growth mentioned above are misleading in important ways. These ultimately derive from their failure to account for ``reproductive excesses'', that is, an excess of juveniles that are more fragile than their adult counterparts \citep{turner1968population}. Reproductive excesses allow selection to be concentrated at a juvenile ``bottleneck'', potentially keeping the density of adults constant even if selection is strong. The reproductive excess concept featured prominently in early debates on the regulation of population density (e.g. \citealt{nicholson_1954}), and is also familiar to evolutionists, particularly in relation to Haldane's ``cost of selection'' \citep{haldane_1957,turner1968population}. Reproductive excesses are also implicit in foundational evolutionary-genetic models like the Wright-Fisher, where each non-overlapping generation involves the production of an infinite number of zygotes, a constant number $N$ of which are sampled to form the next generation of adults. Likewise, in the Moran model, a juvenile is available to replace a dead adult every iteration no matter how rapidly adults are dying, and as a result $N$ remains constant.

Nevertheless, few studies of density-dependent selection have incorporated reproductive excesses. This requires density-regulated population growth in which the excess is finite and, in general, density-dependent. Such models are potentially much more complicated than those with no excess (e.g. logistic) or an infinite excess (e.g. Wright-Fisher). Nei's competitive selection model incorporated a finite reproductive excess to help clarify the ``cost of selection'' \citep{nei1971fertility,nagylaki_1992}, but used an unusual representation of competition based on pairwise interactions defined for at most two different types, and was also restricted to equal fertilities for each type. In more complex models with detailed age structure (as opposed to simply juveniles vs adults), it is usually assumed that the density of a ``critical age-group'' controls the population's response to crowding. The focus is then on the density in the critical age group,  and the role of density dependence is surprisingly similar to that found in models without any age structure \citep[Chap. 4]{charlesworth_1994}. For example, selection proceeds in the direction of increasing equilibrium density in the critical age-group, akin to classical $K$-selection \citep[pp. 148]{charlesworth_1994}. This gives little insight into the role of reproductive excesses in stabilizing adult densities.

We investigate the population ecology of relative fitness using a novel model of density-regulated population growth based on territorial contests, which explicitly incorporates a reproductive excess. We restrict our attention to asexual haploids, since it is then clearer how the success or failure of the constant-$s$ description is tied to the underlying population ecological assumptions. Our starting point is the classic lottery model which was developed by ecologists to study competition driven by territorial contests in reef fishes and plants \citep{sale_77,chesson_1981}. The classic lottery already incorporates a reproductive excess, and fitness involves a product of fertility and juvenile viability akin to standard population genetic models of selection (e.g. \citealt[pp. 185]{crow_1970}). However, the classic lottery also assumes constant $N$. Our first task is to relax the constant-$N$ assumption of the lottery model to create a variable-density lottery (sections ``Model'' and ``Analytical approximation of the variable-density lottery'').

>>>>

In light of the complications of density-dependence, the assignment of density-independent relative fitnesses has been justified as an approximation that holds when selection is weak and $N$ changes slowly (\citealt{kimura1969natural}; \citealt[pp. 277]{ewens_2004}; \citealt[Chap. 4]{charlesworth_1994}). Under these conditions, $s$ is approximately constant in Eq.~\eqref{eq:canonical}, at least for some number of generations. If $s$ depends only on density, not frequency, this approximate constancy can hold over entire selective sweeps \citep[Fig. 9.5]{otto_2011}.

@@

Here we argue that the simplified models of density-regulated growth mentioned above are misleading in their representation of the interplay between selection and density. This ultimately derives from their failure to account for ``reproductive excess'', that is, an excess of juveniles that experience stronger selection than their adult counterparts \citep{turner1968population}. This allows selection to be concentrated at a juvenile ``bottleneck'', potentially keeping the density of adults constant even under strong selection. Reproductive excess featured prominently in early debates about the regulation of population density (e.g. \citealt{nicholson_1954}), and also has a long history in evolutionary theory, particularly related to Haldane's ``cost of selection'' \citep{haldane_1957,turner1968population}. Additionally, reproductive excess is implicit in foundational evolutionary-genetic models like the Wright-Fisher, where each non-overlapping generation involves the production of an infinite number of zygotes, of which a constant number $N$ are sampled to form the next generation of adults. Likewise, in the Moran model, a juvenile is always available to replace a dead adult every iteration no matter how rapidly adults are dying, and as a result $N$ remains constant.

Nevertheless, studies of density-dependent selection rarely incorporate reproductive excess. This requires that we model a finite, density-dependent excess, which is significantly more complicated zero (e.g. logistic) or infinite excess (e.g. Wright-Fisher). Nei's competitive selection model incorporated a finite reproductive excess to help clarify the ``cost of selection'' \citep{nei1971fertility,nagylaki_1992}, but used an unusual representation of competition based on pairwise interactions defined for at most two different genotypes, and was also restricted to equal fertilities for each type. In models with detailed age structure, the role of density is usually absorbed into one ``critical age-group'' that mediates the population's response to crowding. The existence of a reproductive excess implies that the critical age-group would have to occur at a pre-reproductive age. As a result, the focus shifts to juvenile density, obscuring the crucial connection between pre-reproductive-age selection, and post-reproductive-age density. For example, the result that selection proceeds in the direction of increasing equilibrium density in the critical age-group \citep[pp. 148]{charlesworth_1994} is a form of the classical $K$-selection ideas discussed above applied to the critical age-group, and says nothing about other age-groups.

We develop a novel model of density-regulated population growth based on territorial contests, which explicitly incorporates a reproductive excess. Our starting point is the classic lottery model which was developed by ecologists to study competition driven by territorial contests in reef fishes and plants \citep{sale_77,chesson_1981}. The classic lottery already incorporates a reproductive excess, and fitness involves a product of fertility and juvenile viability akin to standard population genetic models of selection (e.g. \citealt[pp. 185]{crow_1970}). However, the classic lottery also assumes constant $N$ (and infinite reproductive excess). Our first task is to relax the constant-$N$ assumption of the lottery model to create a variable-density lottery.

We restrict our attention to asexual haploids, since it is then clearer how the success or failure of the constant-$s$ description is tied to the underlying population ecological assumptions.
